---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


# msSPChelpR

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/msSPChelpR)](https://cran.r-project.org/package=msSPChelpR)
[![](https://img.shields.io/badge/dev%20-0.9.0.9000-green.svg)](https://github.com/marianschmidt/msSPChelpR) 
[![DOI](https://zenodo.org/badge/185602745.svg)](https://zenodo.org/badge/latestdoi/185602745)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/last-month/msSPChelpR?color=grey)](https://marianschmidt.github.io/msSPChelpR/)
[![R-CMD-check](https://github.com/marianschmidt/msSPChelpR/workflows/R-CMD-check/badge.svg)](https://github.com/marianschmidt/msSPChelpR/actions)
<!-- badges: end -->

#### What is `msSPChelpR`?

This package is a collection of helper functions for analyzing Second Primary Cancer (SPC) data, 
including functions to reshape data, to calculate patient states and analyze cancer incidence.


## Installation

You can install the released version of msSPChelpR from [CRAN](https://CRAN.R-project.org) with:

```{r, eval = FALSE}
install.packages("msSPChelpR")
```

And the development version from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
remotes::install_github("marianschmidt/msSPChelpR")
```

## How to use the package

An introductory vignette on the workflow of the package can be found [here](https://marianschmidt.github.io/msSPChelpR/articles/introduction.html). See `vignette("introduction")`.

## Citation

In case you want to cite this package, please use `citation('msSPChelpR')` for reference information.
