---
title: "Introduction to the msSPChelpR package - from long dataset to SIR analyses"
author: "Marian Eberl"
date: "13 May 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pat_status_futime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r , include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(msSPChelpR)
```

# Introduction

This vignette explains how to use the functions:

 * `pat_status` to determine patient status at end of follow-up
 * `vital_status` to determine vital status whether patient is alive or dead at end of follow-up
 * `calc_futime` to calculate follow-up time from index event until next event, death or end of follow-up date

# Recommended Workflows

## Calculate follow-up times

It is recommended to run the following steps in the correct order to obtain accurate follow-up time calculations [Step.

 1. [Use the long version dataset](#step-long) to [filter all cases](#step-filter) that are relevant for your analysis. Make sure that: 
   * for each `case_id` the index event (e.g. First Cancer FC) is still included and is the one remaining row in the dataset with the smallest `case_id` (`TUMID3` variable for ZfKD data, and `SEQ_NUM` for SEER data)
   * all `case_id`s might or might not get a countable incident event (e.g. Second Primary Cancer SPC). This event should be the second entry per `case_id` (second smallest `case_id`) if it is to be counted
   * in the long version dataset a `count_var` should indicate whether the countable incident event (SPC) has occured or not. Coded `0` for non-occurence (or not counted event) and `1` for a counted incident event.
   
 2. [Renumber filtered long dataset](#step-renumber): In the filter long dataset, you should run the helper function `msSPChelpR::renumber_time_id_dt()` (or non-data.table variant `msSPChelpR::renumber_time_id()`) that will renumber all events per `case_id` and (if step 1 is fulfilled) will assign each index event with `time_var_new = 1` and each second (possibly countable incidenct event) with `time_var_new = 2`. Any SIR related function will only count the second event, if additionally to `time_var_new = 2` for this row also `count_var = 1` is true.
 
 3. [Reshape dataset](#step-reshape-wide): Run  `msSPChelpR::reshape_wide_dt()` or non-data.table-variant `msSPChelpR::reshape_wide()`, so that datset is transposed to wide format (1 row per `case_id`, creating variables such as `count_var.2`).
 
 4. [Set flag for Second Primary Cancer diagnosis](#step-spc): After filtering and reshaping it is essential to set `p_spc` again. This variable will be used by later steps of the analysis.
 
 5. [Determine patient status](#step-pat-status) at a defined end of follow-up by using the `msSPChelpR::pat_status()` function. This date for end of follow-up must:
   * be in "YYYY-MM-DD" format and is always defined via the `fu_end =` parameter
   * must preceed the end of data collection. E.g. if the last incident events for the dataset you are using are collected at the end of 2014, your `fu_end` must be `fu_end = "2014-12-15"` or earlier.
   
 6. [Calculate follow-up time](#step-calc-futime) for the same dataset by using the `msSPChelpR::calc_futime()` function and the same `fu_end` as for step 4. By standard all functions of the `msSPChelpR` package require follow-up times as numeric years.
 
 7. 
 
# Details on single functions

## `renumber_time_id` function

## `reshape_wide` function

## `pat_status` function

## `vital_status` function

## `calc_futime` function

# Examples

## SEER lung cancer

### Step 1 - Long dataset {#step-long}




### Step 2 - Filter long dataset {#step-filter}

### Step 3 - Renumber `time_id` {#step-renumber}

### Step 4 - Reshape to wide dataset {#step-reshape-wide}

### Step 5 - Recalculate `p_spc` {#step-spc}

```{r, echo=TRUE, eval=FALSE}

wide_sdata_lung <- wide_sdata_lung %>%
  dplyr::mutate(p_spc = case_when(is.na(PRIMSITE.2)     ~ 0,
                         !is.na(PRIMSITE.2)    ~ 1,
                         TRUE ~ NA_real_)) %>%
  sjlabelled::var_labels(p_spc="Patient has developed Second Primary Cancer") %>%
  sjlabelled::set_labels(p_spc, labels = c("no SPC" = 0,
                                           "SPC developed" = 1)) %>%
  dplyr::mutate_at(vars("p_spc"), sjlabelled::as_label, keep.labels=TRUE) 

```

### Step 6 - Determine patient status at end of FU {#step-pat-status}

### Step 7 - Calculate FU time {#step-calc-futime}

## ZfKD melanoma

### Step 1 - Long dataset

### Step 2 - Filter long dataset

### Step 3 - Renumber `time_id`

### Step 4 - Reshape to wide dataset

### Step 5 - Determine patient status at end of FU

### Step 6 - Calculate FU tim


## Built with

```{r}
sessionInfo()
```
